---
title: "Reproducible Research - Course Project 1"
author: "Umesh Moolchandani"
date: "01/04/2022"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 1 - Loading and preprocessing the data

Code for unzipping, reading in and processing of the data


```{r, echo=TRUE}
setwd("D:/Personal/Coursera/Data Science Specialization/Reproducible Research/RepResearch_Project1")
unzip("./repdata_data_activity.zip")
activityData <- read.csv("./activity.csv")
```

Initial exploration of the data

```{r}
dim(activityData)
names(activityData)
head(activityData)
str(activityData)
#total number of missing data
sum(is.na(activityData$steps))/dim(activityData)[[1]]
#transforming the date column into date format using lubridate
library(lubridate)
activityData$date<-ymd(activityData$date)
length(unique(activityData$date))
```

## Step 2 - Histogram of the total number of steps taken each day

```{r, echo = TRUE}
library(ggplot2)
Q2<-data.frame(tapply(activityData$steps,activityData$date,sum,na.rm=TRUE))
Q2$date<-rownames(Q2)
rownames(Q2)<-NULL
names(Q2)[[1]]<-"Total Steps"
png("plot_1_1.png")
#Bar Chart for Total Steps by date
ggplot(Q2,aes(y=Q2$`Total Steps`,x=Q2$date))+geom_bar(stat="identity") + ylab("Total Steps")+xlab("Date")+ggtitle("Total Steps by date")
dev.off()
ggplot(Q2,aes(y=Q2$`Total Steps`,x=Q2$date))+geom_bar(stat="Identity") + ylab("Total Steps")+xlab("Date")+ggtitle("Total Steps by date")
#Total Steps Histogram
qplot(Q2$`Total Steps`,geom="histogram",xlab="Total Steps",ylab="Counts",main="Total Steps Historgram")
png("plot_1_2.png")
qplot(Q2$`Total Steps`,geom="histogram",xlab="Total Steps",ylab="Counts",main="Total Steps Historgram")
dev.off()
```

## Step 3 - Mean and median number of steps taken each day

```{r, echo = TRUE}
library(dplyr)
Q3<-data.frame(round(tapply(activityData$steps,activityData$date,mean,na.rm=TRUE),2))
Q3$date<-rownames(Q3)
rownames(Q3)<-NULL
names(Q3)[[1]]<-"Mean Steps"
temp<-activityData%>%select(date,steps) %>% group_by(date) %>% summarise(median(steps))
names(temp)[[2]]<-"Median Steps"
Q3$median<-temp$`Median Steps`
Q3<-Q3 %>% select(date,`Mean Steps`,median)
```

## Step 4 - Time series plot of the average number of steps taken

```{r, echo = TRUE}
Q4<-Q3
Q4$date<-as.Date(Q4$date,format="%Y-%m-%d")
ggplot(Q4,aes(x=Q4$date,y=Q4$`Mean Steps`))+geom_bar(stat="identity")+scale_x_date()+ylab("Mean Steps Every day")+xlab("Date")+ggtitle("Mean Steps by Date")
png("plot_4.png")
ggplot(Q4,aes(x=Q4$date,y=Q4$`Mean Steps`))+geom_bar(stat="identity")+scale_x_date()+ylab("Mean Steps Every day")+xlab("Date")+ggtitle("Mean Steps by Date")
dev.off()
```

## Step 5 - The 5-minute interval that, on average, contains the maximum number of steps

```{r, echo = TRUE}
#This is assuming that the words on average means averaging steps by date and interval
activityData$interval<-factor(activityData$interval)
Q5<-aggregate(data=activityData,steps~date+interval,FUN="mean")
Q5<-aggregate(data=Q5,steps~interval,FUN="max")
```


## Step 6

I’ve used the mean/mode value substitution method to update missing values, i.e., using the mean values to substitute the missing values in the original data set.

```{r, echo = TRUE}
Q6<-activityData
Q6$Missing<-is.na(Q6$steps)
Q6<-aggregate(data=Q6,Missing~date+interval,FUN="sum")
Q6.1<-data.frame(tapply(Q6$Missing,Q6$date,sum))
Q6.1$date<-rownames(Q6.1)
rownames(Q6.1)<-NULL
names(Q6.1)<-c("Missing","date")
Q6.1$date<-as.Date(Q6.1$date,format="%Y-%m-%d")
Q6.2<-data.frame(tapply(Q6$Missing,Q6$interval,sum))
Q6.2$date<-rownames(Q6.2)
rownames(Q6.2)<-NULL
names(Q6.2)<-c("Missing","Interval")
par(mfrow=c(1,2))
plot(y=Q6.1$Missing,x=Q6.1$date,main="Missing Value Distribution by Date")
plot(y=Q6.2$Missing,x=Q6.2$Interval,main="Missing Value Distribution by Interval")
table(activityData$date)
```

From the plot, it is evident that the missing values have a very distinct pattern. For every interval, there are consistently 8 missing values. For the date, there are consistently 288 missing values. And in total, there are 8 dates that have missing value. We don't exactly know the cause for these missing values but there's a pattern. For that matter, we can see that the mean value imputation is appropriate.

It is evident that every date has 288 data points. It means that the 8 dates have no data points at all what so ever. We can refine the analysis by looking at these missing values depending on their Weekday and interval parameters to match with the average 

```{r, echo = TRUE}
#Dates that have missing values 
library(lubridate)
Q6.3<-as.data.frame(Q6.1) %>% select(date,Missing) %>% arrange(desc(Missing))
Q6.3<-Q6.3[which(Q6.3$Missing!=0),]
Q6.3$Weekday<-wday(Q6.3$date,label=TRUE)
Q6.4<-activityData
Q6.4$weekday<-wday(Q6.4$date,label=TRUE)
#Finding the mean of steps every monday, and every interval
Q6.5<-aggregate(data=Q6.4,steps~interval+weekday,FUN="mean",na.rm=TRUE)
#Merge the pre-imputation table Q6.4 table with the average table Q6.5
Q6.6<-merge(x=Q6.4,y=Q6.5,by.x=c("interval","weekday"),by.y=c("interval","weekday"),all.x=TRUE)
#Conditionally replacing the steps.x column NA value with the values from steps.y column value 
Q6.6$Steps.Updated<-0
for (i in 1:dim(Q6.6)[[1]]){
if(is.na(Q6.6[i,3])){Q6.6[i,6]=Q6.6[i,5]}
else {Q6.6[i,6]=Q6.6[i,3]}
}
#Now simplify the imputed analytical data frame
Q6.6 <-Q6.6  %>% select(date,weekday,interval,Steps.Updated)
names(Q6.6)[[4]]<-"Steps"
```

## Step 7 - Histogram of the total number of steps taken each day after missing values are imputed

```{r, echo = TRUE}
png("plot_7.png")
qplot(Q6.6$Steps,geom="histogram",main="Total steps taken histogram post imputation",xlab="Steps",ylab="Count")
dev.off()
qplot(Q6.6$Steps,geom="histogram",main="Total steps taken histogram post imputation",xlab="Steps",ylab="Count")
```

## Step 8 - Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

```{r, echo = TRUE}
Q8<-Q6.6
levels(Q8$weekday)<-c(1,2,3,4,5,6,7)
Q8$WDWE<-Q8$weekday %in% c(1,2,3,4,5)
Q8.1<-aggregate(data=Q8,Steps~interval+WDWE,mean,na.rm=TRUE)
Q8.1$WDWE<-as.factor(Q8.1$WDWE)
levels(Q8.1$WDWE)<-c("Weekend","Weekday")
png("plot_8_1.png")
ggplot(data=Q8.1,aes(y=Steps,x=interval,group=1,color=WDWE))+geom_line() +scale_x_discrete(breaks = seq(0, 2500, by = 300))+ylab("Mean Steps")+xlab("Intervals")+ggtitle("Mean steps across intervals by Weekend and Weekday")
dev.off()
ggplot(data=Q8.1,aes(y=Steps,x=interval,group=1,color=WDWE))+geom_line() +scale_x_discrete(breaks = seq(0, 2500, by = 300))+ylab("Mean Steps")+xlab("Intervals")+ggtitle("Mean steps across intervals by Weekend and Weekday")
#Producing the panel plot
Q8.1$interval<-as.numeric(as.character(Q8.1$interval))
library(lattice)
xyplot(data=Q8.1,Steps~interval|WDWE, grid = TRUE, type = c("p", "smooth"), lwd = 4,panel = panel.smoothScatter)
library(hexbin)
hexbinplot(data=Q8.1,Steps~interval|WDWE, aspect = 1, bins=50)
png("plot_8_2.png")
xyplot(data=Q8.1,Steps~interval|WDWE, grid = TRUE, type = c("p", "smooth"), lwd = 4,panel = panel.smoothScatter)
dev.off()
png("plot_8_3.png")
hexbinplot(data=Q8.1,Steps~interval|WDWE, aspect = 1, bins=50)
dev.off()
```

